Fast and scalable Gaussian process modeling of stellar variability \vspace{-3\baselineskip}

### 2016-12-26 paper/ms.tex
edits for SHO

\eq{genrp-kernel-complex} can also be expressed as
\begin{eqnarray}\eqlabel{genrp-kernel}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
    \sum_{j=1}^J &&\left[
    a_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\cos\left(d_j\,\tau_{nm}\right)
        \right.\nonumber\\
    &&+ \left.
    b_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\sin\left(d_j\,\tau_{nm}\right)
\right] \quad.
\end{eqnarray}
The Fourier transform of this covariance function is the power spectral
density (PSD) of the model and it is given by
\begin{eqnarray}
S(\omega) = \sum_{j=1}^p \sqrt{\frac{2}{\pi}}
\frac{(a_j\,c_j+b_j\,d_j)\,({c_j}^2+{d_j}^2)+(a_j\,c_j-b_j\,d_j)\,\omega^2}
{\omega^4+2\,({c_j}^2-{d_j}^2)\,\omega^2+({c_j}^2+{d_j}^2)^2}\quad.
\end{eqnarray}
The physical interpretation of this model isn't immediately obvious and we
will return to a more general discussion of the physical intuition in a moment
but we can start with one useful special case of this model.
If we set the imaginary amplitude $b_j$ for some component $j$ to zero, that
term of \eq{genrp-kernel} becomes
\begin{eqnarray}
k_j(\tau_{nm}) =
    a_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\cos\left(d_j\,\tau_{nm}\right)
\end{eqnarray}
and the PSD for the this component is
\begin{eqnarray}
S_j(\omega) = \frac{1}{\sqrt{2\,\pi}}\,\frac{a_j}{c_j}\,\left[
    \frac{1}{1+\left(\frac{\omega-d_j}{c_j}\right)^2} +
    \frac{1}{1+\left(\frac{\omega+d_j}{c_j}\right)^2}
\right] \quad.
\end{eqnarray}
This PSD is the sum of two Lorentzian or Cauchy distributions with width
$c_j$ centered on $\omega = \pm d_j$.
This model can be interpreted intuitively as a quasiperiodic oscillator with
amplitude $A_j = a_j$, quality factor $Q_j = c_j$, and period $P_j = 2\,\pi /
d_j$.

%The general ACF is given by
%\begin{equation}
%k(\tau) = a_r e^{-b_r \tau} \cos{(b_r \tau)} + a_i e^{-b_r \tau} \sin{(b_r \tau)}.
%\end{equation}
%which has a power spectrum given by
%\begin{equation}
%P(\omega) = P_0 \frac{(a_ib_i+a_rb_r)(b_i^2+b_r^2)+(a_rb_r-a_ib_i)\omega^2}{(\omega^2-(b_i^2+b_r^2))^2 + 4b_r^2\omega^2}.
%\end{equation}
%The power spectrum should be non-negative since it is the absolute square of the Fourier transform of the noise.
%Note that the denominator is always non-negative, so the power spectrum is non-negative if $a_ib_i+a_rb_r \ge 0$ and
%$a_rb_r-a_ib_i \ge 0$.

\subsection{Relation to stochastically-driven, damped simple harmonic oscillator}

Intrinsic stellar variability consists primarily of random fluctuations driven by
convection, oscillations excited by convection, and rotational variability.  These first
two types of variability may be well-described by a model of a damped simple-harmonic oscillator
driven by stochastic variability \citep{1990ApJ...364..699A}.  The governing differential equation is
\begin{equation}
y^{\prime\prime}+ \frac{\omega_0}{Q} y^\prime + \omega_0^2 y = \epsilon,
\end{equation}
where $y(t)$ is a displacement, $^{\prime}$ represents a time derivative,
$\omega_0$ is the frequency of an undamped oscillator, $Q$ is the quality factor
of the oscillation, and $\epsilon(t)$ is a stochastic driving force.

In the limit of an infinite time series and white random forcing, the noise-free power
spectrum of this equation is given by:
\begin{equation}
P(\omega) = P_0 \sqrt{\frac{2}{\pi}} \frac{\omega_0^4}{(\omega^2-\omega_0^2)^2 + \omega_0^2\omega^2/Q^2},
\end{equation}
where $P_0$ is a normalization constant.

>>>>

\eq{genrp-kernel-complex} can also be expressed as
\begin{eqnarray}\eqlabel{genrp-kernel}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
    \sum_{j=1}^J &&\left[
    a_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\cos\left(d_j\,\tau_{nm}\right)
        \right.\nonumber\\
    &&+ \left.
    b_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\sin\left(d_j\,\tau_{nm}\right)
\right] \quad.
\end{eqnarray}
The Fourier transform of this covariance function is the power spectral
density (PSD) of the model and it is given by
\begin{eqnarray}
S(\omega) = \sum_{j=1}^p \sqrt{\frac{2}{\pi}}
\frac{(a_j\,c_j+b_j\,d_j)\,({c_j}^2+{d_j}^2)+(a_j\,c_j-b_j\,d_j)\,\omega^2}
{\omega^4+2\,({c_j}^2-{d_j}^2)\,\omega^2+({c_j}^2+{d_j}^2)^2}\quad.
\end{eqnarray}

The physical interpretation of this model isn't immediately obvious and we
will return to a more general discussion of the physical intuition in a moment
but we can start with one useful special case of this model.
If we set the imaginary amplitude $b_j$ for some component $j$ to zero, that
term of \eq{genrp-kernel} becomes
\begin{eqnarray}
k_j(\tau_{nm}) =
    a_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\cos\left(d_j\,\tau_{nm}\right)
\end{eqnarray}
and the PSD for the this component is
\begin{eqnarray}
S_j(\omega) = \frac{1}{\sqrt{2\,\pi}}\,\frac{a_j}{c_j}\,\left[
    \frac{1}{1+\left(\frac{\omega-d_j}{c_j}\right)^2} +
    \frac{1}{1+\left(\frac{\omega+d_j}{c_j}\right)^2}
\right] \quad.
\end{eqnarray}
This PSD is the sum of two Lorentzian or Cauchy distributions with width
$c_j$ centered on $\omega = \pm d_j$.
This model can be interpreted intuitively as a quasiperiodic oscillator with
amplitude $A_j = a_j$, quality factor $Q_j = c_j$, and period $P_j = 2\,\pi /
d_j$.

Another special case of the \genrp\ model of great physical interest is a
stochastically driven, damped simple harmonic oscillator.
The differential equation for this system is
\begin{equation}
    \left[\frac{\dd^2}{\dd t^2} + \frac{\omega_0}{Q}\,\frac{\dd}{\dd t}
    + \omega_0^2\right]\, y(t) = \epsilon(t)
\end{equation}
where $\omega_0$ is the frequency of the undamped oscillator, $Q$ is the
quality factor of the oscillator, and $\epsilon(t)$ is a stochastic driving
force.
In the limit of an infinite time series and white random forcing, the PSD of
this equation is given by \citep{Anderson:1990}
\begin{equation}
S(\omega) = \sqrt{\frac{2}{\pi}}\,\frac{S_0\,\omega_0^4}
    {(\omega^2-\omega_0^2)^2 + \omega_0^2\omega^2/Q^2}
\end{equation}
where $S_0$ is a normalization constant.
